---
title: "Introduction to simReg"
date: '`r Sys.Date()`'
output: 
  html_document:
    highlight: null
    toc: yes
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to simReg}
-->
```{r setup, include=FALSE}
library(knitr)
library(simReg)
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', kable(x, output = FALSE)), collapse = '\n')
  asis_output(res)
}
```

The *simReg* package aims to define a consistent framework for simulating regression models - including single level and multilevel models. This will hopefully allow the user to quickly simulate data for a class, project, or even a dissertation.

## Installation
Currently development is happening on github. To install the package, use the *devtools* package:
```{r install, eval=FALSE}
library(devtools)
install_github("lebebr01/simReg")
library(simReg)
```
This should load the *devtools* package, install the *simReg* package, and finally load the *simReg* package. The package has currently not been tested on a Mac machine. I do not anticipate any problems installing on a Mac however.

## Simulate Data
The master function that handles the simulation grunt work is the `sim.reg()` function. As always, you can do `?sim.reg` to pull up the help file for the function. Let's look at a simple single level regression example to get started:

```{r singlelevel}
fixed <- ~1 + act + diff + numCourse + act:numCourse
fixed.param <- c(2, 4, 1, 3.5, 2)
cov.param <- list(mean = c(0, 0, 0), sd = c(4, 3, 3), var.type = c("single", "single", "single"))
n <- 150
errorVar <- 3
err.dist <- "norm"
temp.single <- sim.reg(fixed = fixed, fixed.param = fixed.param, cov.param = cov.param,
n = n, errorVar = errorVar, err.dist = err.dist, data.str = "single")
```


## Bugs/Feature Requests
Lastly, for any bugs or feature requests go to the github repository to create post an issue. I will work to resolve them as quickly as possible. See [simReg github repository](https://github.com/lebebr01/simReg)

