---
title: "Simulation Arguments for `simglm`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tidy Simulation with `simglm`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
library(knitr)
library(simglm)
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', kable(x, output = FALSE)), collapse = '\n')
  asis_output(res)
}
```

The tidy simulation framework uses simulation arguments as the basis for specifying the models to be simulated. The goal of this vignette is to document more thoroughly the various arguments that are possible within each function. It is recommended that new users start with the "Tidy Simulation with `simglm`" vignette prior to working through this vignette.

# Fixed Arguments
Arguments associated with the fixed portion of the model are needed for each fixed variable that needs to be simulated. The fixed variables specified come from the formula argument. Interactions or the intercept are not included in the fixed arguments as these are generated automatically. Let's start with an example.

```{r fixed_sim}
library(simglm)

set.seed(321) 

# To-DO: Add knot variable and debug

sim_arguments <- list(
  formula = y ~ 1 + time + weight + age + treat + (1 + time| id),
  fixed = list(time = list(var_type = 'time'),
               weight = list(var_type = 'continuous', mean = 180, sd = 30),
               age = list(var_type = 'ordinal', levels = 30:60, var_level = 2),
               treat = list(var_type = 'factor', 
                            levels = c('Treatment', 'Control'),
                            var_level = 2)),
  sample_size = list(level1 = 10, level2 = 20)
)

fixed_data <- simulate_fixed(data = NULL, sim_arguments)
head(fixed_data, n = 20)
```

The following example shows the five types of variables that can be generated. These types are specified with the `var_type` argument and can be one of the following five types: 

* 'time': This type of variable is used for a time metric variable
* 'continuous': This is a continuous distribution, default is to use the `rnorm` distribution function for generation.
* 'ordinal': This generates ordinal data and is generated using the `sample` function.
* 'factor': This is a categorical or factor variable specification representing groups or categories.
* 'knot': This type of variable is generated based on another variable and represents a knot category variable.

Each variable type will be explored in more detail.

Finally, another common argument for all fixed variable types is the argument `var_level`. This defaults to `var_level = 1` which would be a variable defined at the level 1 of the model (i.e. unique value for each row in the data). These can be changed to reflect the data level that is desired. For example, `var_level = 2` would repeat values for each level 2 cluster found in the data and `var_level = 3` would do the same for each level 3 cluster. Therefore for variables that are at level 2 or level 3, there will be fewer unique values in the data compared to level 1 variables.

## Time Variable
For time variables used in longitudinal or repeated measures designs, the default metric is 0 to level1 sample size minus 1. This can be seen above in the output. To change the time variable metric the `time_levels` argument can be used. A vector of values to specify for the time variable can be given directly. For example, the following could be passed to alter the metric of the time variable: `time_levels = c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5, 6)` where now the time variable would increment in 0.5 units for the first 7 measurements and 1 unit increments for the last three measurements. This could represent a case when the measurements are collected every 6 months for the first 7 measurements and yearly after that.

Below is the output including the manual time variable specification. The only requirement is that the length of the `time_levels` argument must match the level1 sample size.

```{r time_var}
set.seed(321) 

# To-DO: Add knot variable and debug

sim_arguments <- list(
  formula = y ~ 1 + time + weight + age + treat + (1 + time| id),
  fixed = list(time = list(var_type = 'time',
                           time_levels = c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5, 6)),
               weight = list(var_type = 'continuous', mean = 180, sd = 30),
               age = list(var_type = 'ordinal', levels = 30:60, var_level = 2),
               treat = list(var_type = 'factor', 
                            levels = c('Treatment', 'Control'),
                            var_level = 2)),
  sample_size = list(level1 = 10, level2 = 20)
)

fixed_data <- simulate_fixed(data = NULL, sim_arguments)
head(fixed_data, n = 20)
```

## Continuous Variable
Continuous variables are generating using distribution functions (i.e. `rnorm`). Any distribution function found within R, or user written distribution functions can be used, however the default used in `rnorm`. To change the distribution function used, the argument `dist` can be specified. For example, if the Gamma distribution is desired for the weight variable, the following code would achieve this:

```{r rbeta}
set.seed(321) 

sim_arguments <- list(
  formula = y ~ 1 + time + weight + age + treat + (1 + time| id),
  fixed = list(time = list(var_type = 'time',
                           time_levels = c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5, 6)),
               weight = list(var_type = 'continuous', dist = 'rgamma', 
                             shape = 3),
               age = list(var_type = 'ordinal', levels = 30:60, var_level = 2),
               treat = list(var_type = 'factor', 
                            levels = c('Treatment', 'Control'),
                            var_level = 2)),
  sample_size = list(level1 = 10, level2 = 20)
)

fixed_data <- simulate_fixed(data = NULL, sim_arguments)
head(fixed_data, n = 20)
```

This would be the resulting distribution of the weight variable generated.

```{r dist_beta}
library(ggplot2)

ggplot(fixed_data, aes(x = weight)) + 
  geom_density() + 
  theme_bw()
```


## Ordinal Variable
Ordinal variable specification uses the `sample` function within R to generate ordinal variables that are whole integers. The required argument for these types of variables is `levels`. The `levels` argument takes a range or vector of integer values to be passed to the `sample` function within R. For example, these three specifications for the `levels` argument are valid: `3:60`, `seq(4, 60, 2)`, `c(3, 10, 18, 24, 54, 60)`.

An additional optional argument is `replace`. The `replace` argument specified whether the sampling is done with or without replacement. The default behavior is to do sampling with replacement. If sampling without replacement is desired set `replace = FALSE`. See `sample` for more details on this argument.

Finally, the probability of selecting each value specified to the `levels` argument is also able to be specified through the `prob` argument. If `prob` is specified, it takes a vector of probabilities that must be the same length as the levels argument. The default behavior is for each value specified with `levels` to be equally likely to be sampled.

## Factor Variable
Factor variables are generated similarly to ordinal variables with the `sample` function, however factor variables allow the generation of text or categorical variables in addition to numeric grouping variables. Therefore the only needed argument for factor variables is a vector of numeric or text strings representing the different groups to be generated. For example, both of these specifications are equivalent: `c(1, 2, 3, 4)` and `c('Freshman', 'Sophomore', 'Junior', 'Senior')`. Both of these specifications would generate data for these four groups, the difference is that the text labels will be used when text strings are specified. 

An additional optional argument is `replace`. The `replace` argument specified whether the sampling is done with or without replacement. The default behavior is to do sampling with replacement. If sampling without replacement is desired set `replace = FALSE`. See `sample` for more details on this argument.

Finally, the probability of selecting each value specified to the `levels` argument is also able to be specified through the `prob` argument. If `prob` is specified, it takes a vector of probabilities that must be the same length as the levels argument. The default behavior is for each value specified with `levels` to be equally likely to be sampled.

## Knot Variable
To come...

# Random Error Arguments


# Random Effect Arguments


# Model Fit Arguments


# Missing Data Arguments


# Power Arguments

