---
title: "Spline Simulation with `simglm`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spline Simulation with `simglm`}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(simglm)
library(splines)
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', kable(x, output = FALSE)), collapse = '\n')
  asis_output(res)
}
```

# Spline Simulation

Spline terms can be specified directly in the formula argument for fixed-effect simulation. The supported spline basis functions are `ns()` and `bs()`, both from the `splines` package. These terms are expanded into generated basis columns during simulation while the original variable is kept in the simulated data so the same formula can be used for model fitting.

# Natural Cubic Spline

The example below simulates two fixed variables, `x1` and `x2`, and uses a natural cubic spline for `x2`. The formula uses regular R formula syntax.

```{r ns_fixed}
set.seed(321)

sim_arguments <- list(
  formula = y ~ 1 + x1 + ns(x2, df = 4),
  fixed = list(
    x1 = list(var_type = 'continuous', mean = 0, sd = 1),
    x2 = list(var_type = 'continuous', mean = 0, sd = 1)
  ),
  sample_size = 10
)

simulate_fixed(data = NULL, sim_arguments)
```

The generated design matrix contains one column for the intercept, one column for `x1`, four spline basis columns for `x2`, and the original `x2` variable. The spline basis columns are named `x2_1`, `x2_2`, `x2_3`, and `x2_4`.

# Generating an Outcome

Regression weights can be specified with a named list. This is useful for spline terms because one formula term expands to multiple basis columns. The name used for the spline term should match the formula term.

```{r ns_outcome}
set.seed(321)

sim_arguments <- list(
  formula = y ~ 1 + x1 + ns(x2, df = 4),
  fixed = list(
    x1 = list(var_type = 'continuous', mean = 0, sd = 1),
    x2 = list(var_type = 'continuous', mean = 0, sd = 1)
  ),
  error = list(variance = 1),
  sample_size = 100,
  reg_weights = list(
    `(Intercept)` = 0,
    x1 = 0.25,
    `ns(x2, df = 4)` = c(0.75, 0.5, 0, 0)
  )
)

sim_data <- simulate_fixed(data = NULL, sim_arguments) |>
  simulate_error(sim_arguments) |>
  generate_response(sim_arguments)

head(sim_data)
```

# B-Spline Basis

The same pattern can be used for a B-spline basis with `bs()`.

```{r bs_outcome}
set.seed(321)

sim_arguments_bs <- list(
  formula = y ~ 1 + x1 + bs(x2, df = 4),
  fixed = list(
    x1 = list(var_type = 'continuous', mean = 0, sd = 1),
    x2 = list(var_type = 'continuous', mean = 0, sd = 1)
  ),
  error = list(variance = 1),
  sample_size = 100,
  reg_weights = list(
    `(Intercept)` = 0,
    x1 = 0.25,
    `bs(x2, df = 4)` = c(0.75, 0.5, 0, 0)
  )
)

bs_data <- simulate_fixed(data = NULL, sim_arguments_bs) |>
  simulate_error(sim_arguments_bs) |>
  generate_response(sim_arguments_bs)

head(bs_data)
```

# Model Fitting

Spline models can be fit with the existing `model_fit` argument. The original variable is retained in the simulated data, so the fitted model can use the same formula syntax with `ns()` or `bs()`.

```{r ns_model_fit}
set.seed(321)

sim_arguments_fit <- list(
  formula = y ~ 1 + x1 + ns(x2, df = 4),
  fixed = list(
    x1 = list(var_type = 'continuous', mean = 0, sd = 1),
    x2 = list(var_type = 'continuous', mean = 0, sd = 1)
  ),
  error = list(variance = 1),
  sample_size = 100,
  reg_weights = list(
    `(Intercept)` = 0,
    x1 = 0.25,
    `ns(x2, df = 4)` = c(0.75, 0.5, 0, 0)
  ),
  model_fit = list(model_function = 'lm')
)

fit <- simulate_fixed(data = NULL, sim_arguments_fit) |>
  simulate_error(sim_arguments_fit) |>
  generate_response(sim_arguments_fit) |>
  model_fit(sim_arguments_fit)

broom::tidy(fit)
```

# Power Simulation

Spline models can also be used with the existing power simulation workflow. Each
spline basis coefficient is returned as its own model term in the fitted model,
for example `ns(x2, df = 4)1` through `ns(x2, df = 4)4`. Alternative power
thresholds should be named by the fitted model terms returned by `broom::tidy()`.

```{r ns_power}
set.seed(321)

sim_arguments_power <- list(
  formula = y ~ 1 + x1 + ns(x2, df = 4),
  fixed = list(
    x1 = list(var_type = 'continuous', mean = 0, sd = 1),
    x2 = list(var_type = 'continuous', mean = 0, sd = 1)
  ),
  error = list(variance = 1),
  sample_size = 80,
  reg_weights = list(
    `(Intercept)` = 0,
    x1 = 0.25,
    `ns(x2, df = 4)` = c(0.75, 0.5, 0, 0)
  ),
  model_fit = list(model_function = 'lm'),
  power = list(
    thresholds = list(
      x1 = 0.25,
      `ns(x2, df = 4)1` = c(0.5, 0.75),
      `ns(x2, df = 4)2` = 0.5
    )
  ),
  replications = 10,
  extract_coefficients = TRUE
)

power_out <- replicate_simulation(sim_arguments_power)

compute_statistics(
  power_out,
  sim_arguments_power,
  type_1_error = FALSE,
  alternative_power = TRUE
)
```

Random effects can be included with spline fixed effects using the existing random-effect syntax. For example, `y ~ 1 + ns(time, df = 4) + (1 | id)` adds a random intercept while keeping the spline as a fixed-effect term.
